Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-05-23 paper-new/categorical-model.tex
several fixes to paper

Let $c : A \rel A'$ and $f : A \to A'$ in a reflexive graph category
with composition of relations. We say that $c$ is \emph{universally
left-representable by} $f$ if for any $c_l : A_l \rel A$ and $c_r : A'
\rel A_r$ we have $f \ltsq{cc_r}{c_r} \id$ and $\id \ltsq{c_l}{c_lc}
f$. Dually, let $d : B \rel B'$ and $g : B' \to B$.  We say that $d$
is \emph{universally right-representable by} $g$ if for any $d_l : B_l
\rel B$ and $d_r : B' \rel B_r$ we have $\id \ltsq{d_ld}{d_l} \phi$
and $\phi \ltsq{d_r}{dd_r}\id$. In a reflexive graph category these
are stronger than the definitions of left/right representability since
we can pick $c_l,c_r,d_l,d_r$ to be the reflexive relations. In a
double category these are equivalent, but the equivalence uses
horizontal pasting.

Additionally, while in the presence of horizontal pasting
compositionality of representable morphisms is automatic, without this
principle we must require it explicitly. Finally, we explicitly add in
a requirement that the type constructors are functorial in the

In summary, an extensional model consists of:
\begin{enumerate}
  \item A locally thin reflexive graph internal to CBPV models.
  \item Composition of value and computation relations that form a category with the reflexive relations as identity. Call these categories $\mathcal V_r,\mathcal E_r$
  \item Identity-on-objects functors $\upf : \mathcal V_r \to \mathcal V_f$ and $\dnf : \mathcal E_r^{op} \to \mathcal E_f$ such that every $\upf c$ universally left-represents $c$ and every $\dnf d$ universally represents $d$.
  \item The CBPV connectives $U,F,\times,\to$ are all \emph{covariant} functorial on relations up to equivalence: $U(dd') \equidyn U(d)U(d')$ etc.\footnote{the reflexive graph structure already requires that these functors preserve identity relations}
    %% \begin{itemize}
    %% \item $U(dd') \equidyn U(d)U(d')$
    %% \item $F(cc') \equidyn F(c)F(c')$
    %% \item $(cc') \to (dd') \equidyn (c \to d)(c' \to d')$
    %% \item $(c_1c_1') \times (c_2c_2') \equidyn (c_1 \times c_2)(c_1'\times c_2')$
    %% \end{itemize}
    where $c \equidyn c'$ means $\id \ltsq{c}{c'}\id$ and $\id \ltsq{c'}{c} \id$.
  \item A natural transformation $\mho : 1 \Rightarrow U$ such that
    $\mho \circ ! \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$
  \item Distinguished value type $\nat$ with morphisms $z : \mathcal
    V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
  \item Distinguished value types $D$ with distinguished relations
    $\iarr{}: U(D \to F D) \rel D$ and $\inat : \nat \rel D$ and $\itimes : D \times D \rel D$
    each satisfying the retraction property $\dnc {\injarr{}}F(\upc{\injarr{}}) \equidyn \id$.
\end{enumerate}

This mathematical structure is sufficient to interpret the theory of
gradual typing in Section~\ref{sec:GTLC}.
\begin{definition}
  Given any extensional gradual typing model, we interpret
  \begin{enumerate}
  \item Each type $A$ as a value type, interpreting the base types and
    $\times$ as their semantic analogues and $\sem {A \ra A'}$ as $U(\sem{A} \to
    F(\sem{A'}))$.
  \item Type precision derivations $c : A \ltdyn A'$ are
    interpreted as relation morphisms $\sem{c} : \sem{A} \rel \sem{A'}$ in the obvious
    way. Every equivalence axiom $c \equiv c'$ implies that $\sem{c} \equidyn \sem{c'}$.
  \item Every term $\Gamma \vdash M : A$ is interpreted as a morphism
    $\sem{M} : \mathcal V_f (\times\sem{\Gamma},UF\sem{A})$. Upcasts are interpreted as
    $UF(u_{\sem{c}})$ and downcasts as $Ud_{F\sem{c}}$.
  \item If $\Delta \vdash M \ltdyn M' : c$ then $\sem{M}
    \ltsq{\times\sem{\Delta}}{\sem{c}} \sem{M'}$ holds.
  \end{enumerate}
\end{definition}

@@

An intensional model of gradual typing is defined similarly to an extensional model,
with a few key differences that will be discussed below.
%
The starting point is similar to that of the extensional model.
This time, however, since we are working intensionally, the semantic denotation of
term precision \emph{is} transitive, so we \emph{do} have a horizontal composition
operation on squares. Compare this to the extensional case, where we could only
compose \emph{relations} horizontally, not squares.
% What this means is that we can define a functor for composition of value relations
% and squares, and a functor for composition of computation relations and squares.
We can specify this compactly as a category internal to the category of CBPV models
and lax morphisms, where we require that the reflexivity, source, and target morphisms
are strict.
In particular, as in the extensional case, there is a CBPV model of ``objects''
$\mathcal M_f$ and a CBPV model of ``arrows'' $\mathcal M_{sq}$. 
There are strict CBPV morphisms
$r : \mathcal M_f \to \mathcal M_{sq}$ and $s, t : \mathcal M_{sq} \to \mathcal M_f$,
just as before.
%
But now, we also have a CBPV morphism $m$ from the pullback 
$\mathcal M_{sq} \times_{s = t} \mathcal M_{sq}$ to $M_{sq}$, i.e., ``composition of arrows".
In particular, this consists of a functor 
$m_{\mathcal{V}} : \vsq \times_{\sv = \tv} \vsq \to \vsq$ for composition of value
relations/squares, and a functor $m_{\mathcal{E}} : \esq \times_{\se = \te} \esq \to \esq$
for composition of computation relations/squares.
Furthermore, $s \circ m = s \circ \pi_1$ and $t \circ m = t \circ \pi_2$.

As in the extensional model, we also require the existence of a natural transformation
$\mho : 1 \Rightarrow U$ such that $\mho \circ ! \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$,
and a distinguished value type $\nat$ with morphisms $z : \mathcal V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
% Lastly, we require the existence of a morphism $\mho_B \in \vf(\Gamma, UB)$ for all $B$
% satisfying the same ordering and commutativity properties required in the extensional model.

@@

@@

The analogous property should also hold for computation relations and perturbations.
The requirements are summarized below:
%
\begin{definition}\label{def:step-2-model}
  A \emph{step-2} model of intensional gradual typing consists of all the data of a step-1 model plus:
  \begin{enumerate}
    \item For each value type $A$, there is a monoid $P_A$ and a homomorphism of monoids
    $\ptb_A : P_A \to \{ f \in \vf(A,A) \mid f \bisim \id \}$.
    \item For each computation type $B$, there is a monoid $\pe_B$ and a homomorphism of monoids
    $\ptb_B : P_B \to \{ g \in \ef(B,B) \mid g \bisim \id \}$.
    \item For all $B$, the distinguished endomorphism $\delta_{UB}^*$ is in $P_{UB}$.
    % \item $\pv_A$ and a monoid homomorphism 
    %   \[ \ptbv_A : \pv_A \to \{ f \in \vf(A,A) \mid f \bisim \id \} \]
    % \item $\pe_B$ and a monoid homomorphism
    %   \[ \ptbe_B : \pe_B \to \{ g \in \ef(B,B) \mid g \bisim \id \} \]
    \item The functors $\times, \timesk$, $\arr, \tok$, $U, \Uk$, $F, \Fk$ preserve perturbations.
    \item The push-pull property holds for all $c : A \rel A'$ and all $d : B \rel B'$.
  \end{enumerate}
\end{definition}

@@

First, let $\mathcal M$ be any double category with a notion of perturbations,
i.e., for any object $X$ in $\mathcal M$ there is a monoid $P_X$ with a monoid
homomorphism into the endomorphisms on $X$.
%
\begin{definition}\label{def:quasi-left-representable}
  Let $R$ be a horizontal morphism in $\mathcal M$ between objects $X$ and $Y$.
  We say that $R$ is quasi-left-representable by a vertical morphism $f$ in
  $\mathcal M(X, Y)$ if there are perturbations $\delle_R \in P_X$ and
  $\delre_R \in P_Y$ such that there is a square
  $\upl : f \ltsq{R}{r(Y)} \delre_R$ and a square 
  $\upr : \delle_R \ltsq{r(X)}{R}$
%
%   \begin{center}
%     \begin{tabular}{ m{7em} m{7em} } 
%       % UpL
%       \begin{tikzcd}[ampersand replacement=\&]
%         X \& {Y} \\
%         {Y} \& {Y}
%         \arrow["f"', from=1-1, to=2-1]
%         \arrow["\delta_R^{r,e}", from=1-2, to=2-2]
%         \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
%         \arrow[from=2-1, to=2-2, Rightarrow, no head]
%       \end{tikzcd}
%       &
%       % UpR
%       \begin{tikzcd}[ampersand replacement=\&]
%         X \& {X} \\
%         {X} \& {Y}
%         \arrow["\delta_R^{l,e}"', from=1-1, to=2-1]
%         \arrow["f", from=1-2, to=2-2]
%         \arrow[from=1-1, to=1-2, Rightarrow, no head]
%         \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
%       \end{tikzcd}
%     \end{tabular}
%   \end{center}
% %
%   We call the first square $\upl$ and the second square $\upr$. 
%
\end{definition}
%
\begin{definition}\label{def:quasi-right-representable}
  Let $R$ be a horizontal morphism between $X$ and $Y$. We say that $R$ is
  \emph{quasi-right-representable by} $f \in \mathcal M(Y, X)$
  if there exist perturbations $\dellp_R \in P_X$ and
  $\delrp_R \in P_Y$ such that we have a square 
  $\dnr : \dellp_R \ltsq{R}{r(X)} f$
  and a square $\dnl : f \ltsq{r(Y)}{R} \delrp_R$.
  %
  % \begin{center}
  %   \begin{tabular}{ m{7em} m{7em} } 
  %     % DnR
  %     \begin{tikzcd}[ampersand replacement=\&]
  %       {X} \& {Y} \\
  %       {X} \& {X}
  %       \arrow["\delta_R^{l,p}"', from=1-1, to=2-1]
  %       \arrow["f", from=1-2, to=2-2]
  %       \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
  %       \arrow[from=2-1, to=2-2, Rightarrow, no head]
  %     \end{tikzcd}
  %     &
  %     % DnL
  %     \begin{tikzcd}[ampersand replacement=\&]
  %       {Y} \& {Y} \\
  %       {X} \& {Y}
  %       \arrow["f"', from=1-1, to=2-1]
  %       \arrow["\delta_R^{r,p}", from=1-2, to=2-2]
  %       \arrow[from=1-1, to=1-2, Rightarrow, no head]
  %       \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
  %     \end{tikzcd}
  %   \end{tabular}
  % \end{center}
  % %
  % We call the first square $\dnr$ and the second square $\dnl$.
  %
  \end{definition}

@@

Lastly, we require that the model satisfy a weak version of functoriality for 
the CBPV connectives $U,F,\times,\to$. 
First, we will need a definition:
%
\begin{definition}[quasi-order-equivalence]\label{def:quasi-order-equivalent}
  Let $c, c' : A \rel A'$. We say that $c$ and $c'$ are \emph{quasi-order-equivalent},
  written $c \qordeq c'$, if there exist perturbations $\delta^l_1, \delta^l_2 \in \pv_A$ and 
  $\delta^r_1, \delta^r_2 \in \pv_{A'}$ such that there is a square
  $\delta^l_1 \ltsq{c}{c'} \delta^r_1$ and a square $\delta^l_2 \ltsq{c'}{c} \delta^r_2$.
%
  % \begin{center}
  %   \begin{tabular}{ m{9em} m{9em} } 
  %     \begin{tikzcd}[ampersand replacement=\&]
  %       A \& {A'} \\
  %       A \& {A'}
  %       \arrow["\delta^l_1"', from=1-1, to=2-1]
  %       \arrow["\delta^r_1", from=1-2, to=2-2]
  %       \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
  %       \arrow["c'"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
  %     \end{tikzcd}
  %     &
  %     \begin{tikzcd}[ampersand replacement=\&]
  %       A \& {A'} \\
  %       A \& {A'}
  %       \arrow["\delta^l_2"', from=1-1, to=2-1]
  %       \arrow["\delta^r_2", from=1-2, to=2-2]
  %       \arrow["c'", "\shortmid"{marking}, no head, from=1-1, to=1-2]
  %       \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
  %     \end{tikzcd}
  %   \end{tabular}
  % \end{center}
%
  We make the analogous definition for computation relations $d, d' : B \rel B'$.
\end{definition}
%
We require that the CBPV connectives $U,F,\times,\to$ are \emph{quasi-functorial} on relations,
which we specify as follows:
\begin{itemize}
  \item $U(d \comp d') \qordeq U(d)U(d')$
  \item $F(c \comp c'') \qordeq F(c)F(c')$
  \item $(cc') \to (dd') \qordeq (c \to d)(c' \to d')$
  \item $(c_1c_1') \times (c_2c_2') \qordeq (c_1 \times c_2)(c_1'\times c_2')$
\end{itemize}

We summarize the requirements of a step-3 model below:
%
\begin{definition}\label{def:step-3-model}
  A \emph{step-3 intensional model} consists
  of all the data of a step-2 intensional model, such that additionally:
  \begin{enumerate}
    \item There are functors $\upf : \ve \to \vf$ and $\dnf : \ee^{op} \to \ef$ % TODO: image is thin?
    \item Every value edge $c : A \rel A'$ is quasi-left-representable by $\upf(c)$ and
    every computation edge $d : B \rel B'$ is quasi-right-representable by $\dnf(d)$.
    \item The CBPV connectives $U,F,\times,\to$ are quasi-functorial on relations.
  \end{enumerate}
\end{definition}

@@

In the previous section, we have given the definition of an intensional model
of gradual typing as a series of steps with each definition building on the previous one.
%
Here, we discuss how to construct an extensional model from an intensional model with dyn.
We do so in several phases, beginning with a step-1 intensional model with dyn
and ending with an extensional model.
Moreover, this construction is \emph{modular}, in that each phase of the
construction does not depend on the details of the previous ones.
% However, this process cannot proceed in isolation: some phases require
% additional inputs. We will make clear what data must be supplied to each phase.

@@

Now suppose we have a step-2 intensional model $\mathcal{M}$.
We claim that we can construct a \hyperref[def:step-3-model]{step-3 intensional model} $\mathcal{M'}$.
In the construction, the objects and morphisms are the same as those of $\mathcal M$,
and a value relation between $A$ and $A'$ consists of a triple $(c, \rho^L_c, \rho^R_{Fc})$
where $c : A \rel A'$ is a relation in $\mathcal{M}$,
$\rho^L_c$ is a quasi-\emph{left}-representation for $c$ 
(i.e., an embedding $e_c$, perturbations $\delre_c \in P_{A'}$ and $\delle_c \in P_A$, 
and the two relevant squares), and similarly $\rho^R_{Fc}$ is
a quasi-\emph{right}-representation for $Fc$.
Computation relations are triples $(d, \rho^R_d, \rho^L_{Ud})$ where $d : B \rel B'$,
$\rho^R_d$ is a quasi-right-representation for $d$ and $\rho^L{Ud}$ is a quasi-left-representation for $Ud$.
The value and computation squares are the same as those of $\mathcal{M}$.
We then define composition of relations and the action of the functors $F$, $U$, $\times$, and $\arr$.

>>>>

Let $c : A \rel A'$ and $f : A \to A'$ in a reflexive graph category
with composition of relations. We say that $c$ is \emph{universally
left-representable by} $f$ if for any $c_l : A_l \rel A$ and $c_r : A'
\rel A_r$ we have $f \ltsq{cc_r}{c_r} \id$ and $\id \ltsq{c_l}{c_lc}
f$. Dually, let $d : B \rel B'$ and $\phi : B' \to B$.  We say that $d$
is \emph{universally right-representable by} $\phi$ if for any $d_l : B_l
\rel B$ and $d_r : B' \rel B_r$ we have $\id \ltsq{d_ld}{d_l} \phi$
and $\phi \ltsq{d_r}{dd_r}\id$. In a reflexive graph category these
are stronger than the definitions of left/right representability since
we can pick $c_l,c_r,d_l,d_r$ to be the reflexive relations. In a
double category these are equivalent, but the equivalence uses
horizontal pasting.

Additionally, while in the presence of horizontal pasting
compositionality of representable morphisms is automatic, without this
principle we must require it explicitly. Finally, we explicitly add in
a requirement that the type constructors are functorial on relations
up to order equivalence.

In summary, an extensional model consists of:
\begin{enumerate}
  \item A locally thin reflexive graph internal to CBPV models.
  \item Composition of value and computation relations that form a category with the reflexive relations as identity. Call these categories $\mathcal V_r,\mathcal E_r$
  \item Identity-on-objects functors $\upf : \mathcal V_r \to \mathcal V_f$ and $\dnf : \mathcal E_r^{op} \to \mathcal E_f$ such that every $\upf\, c$ universally left-represents $c$ and every $\dnf\, d$ universally right-represents $d$.
  \item The CBPV connectives $U,F,\times,\to$ are all \emph{covariant} functorial on relations up to equivalence: $U(dd') \equidyn U(d)U(d')$ etc.\footnote{the reflexive graph structure already requires that these functors preserve identity relations}
    %% \begin{itemize}
    %% \item $U(dd') \equidyn U(d)U(d')$
    %% \item $F(cc') \equidyn F(c)F(c')$
    %% \item $(cc') \to (dd') \equidyn (c \to d)(c' \to d')$
    %% \item $(c_1c_1') \times (c_2c_2') \equidyn (c_1 \times c_2)(c_1'\times c_2')$
    %% \end{itemize}
    where $c \equidyn c'$ means $\id \ltsq{c}{c'}\id$ and $\id \ltsq{c'}{c} \id$.
  \item A natural transformation $\mho : 1 \Rightarrow U$ such that
    $\mho \circ {!} \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$
  \item Distinguished value type $\nat$ with morphisms $z : \mathcal
    V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
  \item Distinguished value types $D$ with distinguished relations
    $\iarr{}: U(D \to F D) \rel D$ and $\inat : \nat \rel D$ and $\itimes : D \times D \rel D$
    each satisfying the retraction property $\dnc {\iarr{}}F(\upc{\iarr{}}) \equidyn \id$.
\end{enumerate}

This mathematical structure is sufficient to interpret the theory of
gradual typing in Section~\ref{sec:GTLC}.
\begin{definition}
  Given any extensional gradual typing model, we interpret
  \begin{enumerate}
  \item Each type $A$ as a value type, interpreting the base types and
    $\times$ as their semantic analogues and $\sem {A \ra A'}$ as $U(\sem{A} \to
    F(\sem{A'}))$.
  \item Type precision derivations $c : A \ltdyn A'$ are
    interpreted as relation morphisms $\sem{c} : \sem{A} \rel \sem{A'}$ in the obvious
    way. Every equivalence axiom $c \equiv c'$ implies that $\sem{c} \equidyn \sem{c'}$.
  \item Every term $\Gamma \vdash M : A$ is interpreted as a morphism
    $\sem{M} : \mathcal V_f (\times\sem{\Gamma},UF\sem{A})$. Upcasts are interpreted as
    $UF(\upf\, {\sem{c}})$ and downcasts as $U (\dnf\, {(F\sem{c})})$.
  \item If $\Delta \vdash M \ltdyn M' : c$ then $\sem{M}
    \ltsq{\times\sem{\Delta}}{\sem{c}} \sem{M'}$ holds.
  \end{enumerate}
\end{definition}

@@

An intensional model of gradual typing is defined similarly to an extensional model,
with a few key differences that will be discussed below.
To aid in understanding the definition of an intensional model, we break the
definition into five steps, with each one building on the previous.
The last step will be the definition we have in mind when we refer to an intensional model.

% This also facilitates the construction of an intensional model by breaking it down
% into a few modular constructions between each subsequent phase.

%
The starting point is similar to that of the extensional model.
This time, however, since we are working intensionally, the semantic denotation of
term precision \emph{is} transitive, so we \emph{do} have a horizontal composition
operation on squares. Compare this to the extensional case, where we could only
compose \emph{relations} horizontally, not squares.
% What this means is that we can define a functor for composition of value relations
% and squares, and a functor for composition of computation relations and squares.
We can specify this compactly as a category internal to the category of CBPV models
and lax morphisms, where we require that the reflexivity, source, and target morphisms
are strict.
In particular, as in the extensional case, there is a CBPV model of ``objects''
$\mathcal M_f$ and a CBPV model of ``arrows'' $\mathcal M_{sq}$. 
There are strict CBPV morphisms
$r : \mathcal M_f \to \mathcal M_{sq}$ and $s, t : \mathcal M_{sq} \to \mathcal M_f$,
just as before.
%
But now, we also have a CBPV morphism $m$ from the pullback 
$\mathcal M_{sq} \times_{s = t} \mathcal M_{sq}$ to $\mathcal M_{sq}$, i.e., ``composition of arrows".
In particular, this consists of a functor 
$m_{\mathcal{V}} : \vsq \times_{\sv = \tv} \vsq \to \vsq$ for composition of value
relations/squares, and a functor $m_{\mathcal{E}} : \esq \times_{\se = \te} \esq \to \esq$
for composition of computation relations/squares.
Furthermore, $s \circ m = s \circ \pi_1$ and $t \circ m = t \circ \pi_2$.

As in the extensional model, we also require the existence of a natural transformation
$\mho : 1 \Rightarrow U$ such that $\mho \circ {!} \ltsq{r(A)}{r(UB)} f$ for any $f : A \to UB$,
and a distinguished value type $\nat$ with morphisms $z : \mathcal V(1,\nat)$ and $s : \mathcal V(\nat,\nat)$.
% Lastly, we require the existence of a morphism $\mho_B \in \vf(\Gamma, UB)$ for all $B$
% satisfying the same ordering and commutativity properties required in the extensional model.

@@

We call a model satisfying these properties a \emph{step-0 intensional model}.

@@

The analogous property should also hold for computation relations and perturbations.
The requirements are summarized below:
%
\begin{definition}\label{def:step-2-model}
  A \emph{step-2} model of intensional gradual typing consists of all the data of a step-1 model plus:
  \begin{enumerate}
    \item For each value type $A$, there is a monoid $\pv_A$ and a homomorphism of monoids
    $\ptb_A : \pv_A \to \{ f \in \vf(A,A) \mid f \bisim \id \}$.
    \item For each computation type $B$, there is a monoid $\pe_B$ and a homomorphism of monoids
    $\ptb_B : \pe_B \to \{ g \in \ef(B,B) \mid g \bisim \id \}$.
    \item For all $B$, the distinguished endomorphism $\delta_{UB}^*$ is in $\pv_{UB}$.
    % \item $\pv_A$ and a monoid homomorphism 
    %   \[ \ptbv_A : \pv_A \to \{ f \in \vf(A,A) \mid f \bisim \id \} \]
    % \item $\pe_B$ and a monoid homomorphism
    %   \[ \ptbe_B : \pe_B \to \{ g \in \ef(B,B) \mid g \bisim \id \} \]
    \item The functors $\times, \timesk$, $\arr, \tok$, $U, \Uk$, $F, \Fk$ preserve perturbations.
    \item The push-pull property holds for all $c : A \rel A'$ and all $d : B \rel B'$.
  \end{enumerate}
\end{definition}

@@

First, let $\mathcal M$ be any double category with a notion of perturbations,
i.e., for any object $X$ in $\mathcal M$ there is a monoid $P_X$ with a monoid
homomorphism into the endomorphisms on $X$.
%
\begin{definition}\label{def:quasi-left-representable}
  Let $R$ be a horizontal morphism in $\mathcal M$ between objects $X$ and $Y$.
  We say that $R$ is \emph{quasi-left-representable by} a vertical morphism $f$ in
  $\mathcal M(X, Y)$ if there are perturbations $\delle_R \in P_X$ and
  $\delre_R \in P_Y$ such that there is a square
  $\upl : f \ltsq{R}{r(Y)} \delre_R$ and a square 
  $\upr : \delle_R \ltsq{r(X)}{R} f$.
%
%   \begin{center}
%     \begin{tabular}{ m{7em} m{7em} } 
%       % UpL
%       \begin{tikzcd}[ampersand replacement=\&]
%         X \& {Y} \\
%         {Y} \& {Y}
%         \arrow["f"', from=1-1, to=2-1]
%         \arrow["\delta_R^{r,e}", from=1-2, to=2-2]
%         \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
%         \arrow[from=2-1, to=2-2, Rightarrow, no head]
%       \end{tikzcd}
%       &
%       % UpR
%       \begin{tikzcd}[ampersand replacement=\&]
%         X \& {X} \\
%         {X} \& {Y}
%         \arrow["\delta_R^{l,e}"', from=1-1, to=2-1]
%         \arrow["f", from=1-2, to=2-2]
%         \arrow[from=1-1, to=1-2, Rightarrow, no head]
%         \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
%       \end{tikzcd}
%     \end{tabular}
%   \end{center}
% %
%   We call the first square $\upl$ and the second square $\upr$. 
%
\end{definition}
%
\begin{definition}\label{def:quasi-right-representable}
  Let $R$ be a horizontal morphism between $X$ and $Y$. We say that $R$ is
  \emph{quasi-right-representable by} $f \in \mathcal M(Y, X)$
  if there exist perturbations $\dellp_R \in P_X$ and
  $\delrp_R \in P_Y$ such that we have a square 
  $\dnr : \dellp_R \ltsq{R}{r(X)} f$
  and a square $\dnl : f \ltsq{r(Y)}{R} \delrp_R$.
  %
  % \begin{center}
  %   \begin{tabular}{ m{7em} m{7em} } 
  %     % DnR
  %     \begin{tikzcd}[ampersand replacement=\&]
  %       {X} \& {Y} \\
  %       {X} \& {X}
  %       \arrow["\delta_R^{l,p}"', from=1-1, to=2-1]
  %       \arrow["f", from=1-2, to=2-2]
  %       \arrow["R", "\shortmid"{marking}, no head, from=1-1, to=1-2]
  %       \arrow[from=2-1, to=2-2, Rightarrow, no head]
  %     \end{tikzcd}
  %     &
  %     % DnL
  %     \begin{tikzcd}[ampersand replacement=\&]
  %       {Y} \& {Y} \\
  %       {X} \& {Y}
  %       \arrow["f"', from=1-1, to=2-1]
  %       \arrow["\delta_R^{r,p}", from=1-2, to=2-2]
  %       \arrow[from=1-1, to=1-2, Rightarrow, no head]
  %       \arrow["R"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
  %     \end{tikzcd}
  %   \end{tabular}
  % \end{center}
  % %
  % We call the first square $\dnr$ and the second square $\dnl$.
  %
  \end{definition}

@@

Lastly, we require that the model satisfy a weak version of functoriality for 
the CBPV connectives $U,F,\times,\to$. 
First, we will need a definition:
%
\begin{definition}[quasi-order-equivalence]\label{def:quasi-order-equivalent}
  Let $c, c' : A \rel A'$. We say that $c$ and $c'$ are \emph{quasi-order-equivalent},
  written $c \qordeq c'$, if there exist perturbations $\delta^l_1, \delta^l_2 \in \pv_A$ and 
  $\delta^r_1, \delta^r_2 \in \pv_{A'}$ such that there is a square
  $\delta^l_1 \ltsq{c}{c'} \delta^r_1$ and a square $\delta^l_2 \ltsq{c'}{c} \delta^r_2$.
%
  % \begin{center}
  %   \begin{tabular}{ m{9em} m{9em} } 
  %     \begin{tikzcd}[ampersand replacement=\&]
  %       A \& {A'} \\
  %       A \& {A'}
  %       \arrow["\delta^l_1"', from=1-1, to=2-1]
  %       \arrow["\delta^r_1", from=1-2, to=2-2]
  %       \arrow["c", "\shortmid"{marking}, no head, from=1-1, to=1-2]
  %       \arrow["c'"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
  %     \end{tikzcd}
  %     &
  %     \begin{tikzcd}[ampersand replacement=\&]
  %       A \& {A'} \\
  %       A \& {A'}
  %       \arrow["\delta^l_2"', from=1-1, to=2-1]
  %       \arrow["\delta^r_2", from=1-2, to=2-2]
  %       \arrow["c'", "\shortmid"{marking}, no head, from=1-1, to=1-2]
  %       \arrow["c"', "\shortmid"{marking}, no head, from=2-1, to=2-2]
  %     \end{tikzcd}
  %   \end{tabular}
  % \end{center}
%
  We make the analogous definition for computation relations $d, d' : B \rel B'$.
\end{definition}
%
We require that the CBPV connectives $U,F,\times,\to$ are \emph{quasi-functorial} on relations,
which we specify as follows, with $c \comp c'$ denoting the relational composition of $c$ and $c'$:
\begin{itemize}
  \item $U(d \comp d') \qordeq U(d) \comp U(d')$
  \item $F(c \comp c') \qordeq F(c) \comp F(c')$
  \item $(c \comp c') \to (d \comp d') \qordeq (c \to d) \comp (c' \to d')$
  \item $(c_1 \comp c_1') \times (c_2 \comp c_2') \qordeq (c_1 \times c_2) \comp (c_1'\times c_2')$
\end{itemize}

We summarize the requirements of a step-3 model below:
%
\begin{definition}\label{def:step-3-model}
  A \emph{step-3 intensional model} consists
  of all the data of a step-2 intensional model, such that additionally:
  \begin{enumerate}
    \item There are functors $\upf : \vr \to \vf$ and $\dnf : \er^{op} \to \ef$ % TODO: image is thin?
    \item Every value edge $c : A \rel A'$ is quasi-left-representable by $\upf(c)$ and
    every computation edge $d : B \rel B'$ is quasi-right-representable by $\dnf(d)$.
    \item The CBPV connectives $U,F,\times,\to$ are quasi-functorial on relations.
  \end{enumerate}
\end{definition}

@@

In the previous section, we have given the definition of an intensional model
of gradual typing as a series of steps with each definition building on the previous one.
%
Here, we discuss how to construct an extensional model from a step-4 intensional model.
We do so in several phases, beginning with a step-1 intensional model with dyn
and ending with an extensional model.
Moreover, this construction is \emph{modular}, in that each phase of the
construction does not depend on the details of the previous ones.
% However, this process cannot proceed in isolation: some phases require
% additional inputs. We will make clear what data must be supplied to each phase.

@@

Now suppose we have a step-2 intensional model $\mathcal{M}$.
We claim that we can construct a \hyperref[def:step-3-model]{step-3 intensional model} $\mathcal{M'}$.
In the construction, the objects and morphisms are the same as those of $\mathcal M$,
and a value relation between $A$ and $A'$ consists of a triple $(c, \rho^L_c, \rho^R_{Fc})$
where $c : A \rel A'$ is a relation in $\mathcal{M}$,
$\rho^L_c$ is a quasi-\emph{left}-representation for $c$ 
(i.e., an embedding $e_c$, perturbations $\delre_c \in P_{A'}$ and $\delle_c \in P_A$, 
and the two relevant squares), and similarly $\rho^R_{Fc}$ is
a quasi-\emph{right}-representation for $Fc$.
Computation relations are triples $(d, \rho^R_d, \rho^L_{Ud})$ where $d : B \rel B'$,
$\rho^R_d$ is a quasi-right-representation for $d$ and $\rho^L_{Ud}$ is a quasi-left-representation for $Ud$.
The value and computation squares are the same as those of $\mathcal{M}$.
We then define composition of relations and the action of the functors $F$, $U$, $\times$, and $\arr$.
